---
title: "DataSHIELD OMOP Practical Exercises"
subtitle: "Learning to explore and analyze OMOP data with DataSHIELD"
freeze: true
format: 
  html:
    code-copy: true
    toc: true
    toc-depth: 3
execute:
  eval: true
  echo: true
  warning: false
  error: true
---

<a href="#" onclick="event.preventDefault(); fetch('https://raw.githubusercontent.com/isglobal-brge/workshop_DSWB/main/omop_exercises_solution.R').then(response => response.blob()).then(blob => { const url = window.URL.createObjectURL(blob); const a = document.createElement('a'); a.style.display = 'none'; a.href = url; a.download = 'omop_exercises_solution.R'; document.body.appendChild(a); a.click(); window.URL.revokeObjectURL(url); }); return false;" class="btn btn-primary btn-sm">
<i class="bi bi-download"></i> Run the solution yourself!
</a>

# Introduction

This tutorial will guide you through exploring and analyzing OMOP data using DataSHIELD. You'll work with a real dataset from the African Population and Health Research Center (APHRC), discovering what type of study it is through systematic exploration.

## Learning Objectives

By the end of this tutorial, you will learn how to:

- Connect to an OMOP database through DataSHIELD
- Explore available data systematically to understand the study design
- Transform OMOP concepts into analyzable variables
- Perform privacy-preserving statistical analyses
- Build and interpret statistical models with federated data

## Setup

First, let's load the required libraries and configure SSL settings:

```{r setup}
# Load required libraries
library(DSI)
library(DSOpal)
library(dsBaseClient)
library(dsOMOPClient)
library(dsOMOPHelper)

# Configure SSL settings (for self-signed certificates)
library(httr)
httr::set_config(httr::config(ssl_verifypeer = 0L, ssl_verifyhost = 0L))
```

# Exercise 1: Server Connection and Initial Data Exploration

Let's connect to the mystery dataset and see what we can discover!

```{r exercise1}
cat("\n==================================================\n")
cat("EXERCISE 1: Connect to the server and explore the data structure\n")
cat("==================================================\n\n")

# Create a connection to the OMOP DataSHIELD server
builder <- newDSLoginBuilder()
builder$append(server="aphrc",
               url="https://44.201.204.234:9090",
               user="administrator",
               password="password",
               driver = "OpalDriver",
               profile = "omop")

logindata <- builder$build()
conns <- datashield.login(logins=logindata)

# Create a dsOMOPHelper instance
o <- ds.omop.helper(
  connections = conns,
  resource = "omop.study_1", 
  symbol = "data"
)

# Check the initial data summary
cat("Initial data summary:\n")
ds.summary("data")
```

# Exercise 2: Who Are We Studying?

Let's investigate the characteristics of our study population.

```{r exercise2}
cat("\n\n==================================================\n")
cat("EXERCISE 2: Who Are We Studying?\n")
cat("==================================================\n\n")

# First Clue - Gender Distribution
cat("Task 2.1: Gender distribution\n")
ds.table("data$gender_concept_id")
cat("\nObservation: All participants are female! This is our first clue.\n")

# Second Clue - Age Distribution
cat("\nTask 2.2: Age distribution\n")
ds.assign(
  toAssign = "2025 - data$year_of_birth",
  newobj = "age",
  datasources = conns
)
cat("Age statistics:\n")
ds.mean("age", datasources = conns)
ds.quantileMean("age", datasources = conns)
cat("\nObservation: Women of reproductive age (roughly 30-45)\n")

# Sample Size
cat("\nTask 2.3: Sample size\n")
print(ds.dim("data"))
cat("\nWe have 188 female participants of reproductive age. What might they have in common?\n")
```

# Exercise 3: What Was Measured?

Now let's explore what types of data were collected.

```{r exercise3}
cat("\n\n==================================================\n")
cat("EXERCISE 3: What Was Measured?\n")
cat("==================================================\n\n")

# Explore Conditions (Diagnoses)
cat("Task 3.1: Available conditions:\n")
cond_concepts <- o$concepts("condition_occurrence")
print(cond_concepts)
cat("\nReveal: Depression during and after pregnancy! This is a maternal mental health study.\n")

# Explore Measurements
cat("\nTask 3.2: Clinical measurements:\n")
meas_concepts <- o$concepts("measurement")
print(meas_concepts)
cat("\nThe Edinburgh Postnatal Depression Scale - a standard tool for screening maternal depression\n")

# Explore Observations
cat("\nTask 3.3: Social and demographic factors:\n")
obs_concepts <- o$concepts("observation")
print(obs_concepts)
cat("\nKey factors: Employment, marital status, household size, religion, education\n")
cat("These social determinants can influence mental health outcomes!\n")
```

# Exercise 4: Building Your Analysis Dataset

Let's retrieve the variables we need for our analysis.

```{r exercise4}
cat("\n\n==================================================\n")
cat("EXERCISE 4: Building Your Analysis Dataset\n")
cat("==================================================\n\n")

# Retrieve the main clinical measurement
cat("Task 4.1: Retrieving Edinburgh Postnatal Depression Scale...\n")
o$auto(tables="measurement", 
       concepts=4164838,
       columns="value_as_number")

cat("\nData structure after adding Edinburgh scale:\n")
ds.summary("data")
cat("\n✓ Added: edinburgh_postnatal_depression_scale.value_as_number\n")

# Add Social Determinants
cat("\nTask 4.2: Retrieving social determinants...\n")

# Employment status
cat("- Adding employment status...\n")
o$auto(tables="observation",
       concepts=44804285,
       columns="value_as_concept_id")

# Other social factors
cat("- Adding marital status, household size, and religious affiliation...\n")
o$auto(tables="observation",
       concepts=c(4053609, 4075500, 4052017),
       columns="value_as_concept_id")

# Education level
cat("- Checking for education level...\n")
o$auto(tables="observation",
       concepts=42528763,
       columns="value_as_concept_id")

cat("\nCurrent variables in our dataset:\n")
current_vars <- ds.colnames("data")[[1]]
print(current_vars)

# Retrieve the Diagnoses
cat("\nTask 4.3: Retrieving depression diagnoses...\n")
o$auto(tables="condition_occurrence",
       concepts=c(4239471, 37312479),
       columns="condition_occurrence_id")

# Final check
cat("\nFinal data structure:\n")
ds.summary("data")
```

# Exercise 5: The Data Transformation Challenge

The depression diagnoses are stored as IDs. Let's understand why we need to transform them.

```{r exercise5}
cat("\n\n==================================================\n")
cat("EXERCISE 5: The Data Transformation Challenge\n")
cat("==================================================\n\n")

cat("Examining the depression diagnosis columns...\n")
cat("These are stored as condition_occurrence_id values (long ID numbers)\n")
cat("We need to transform these into analyzable boolean (0/1) variables\n\n")

# Let's see what columns we have now
current_cols <- ds.colnames("data")[[1]]
depression_cols <- grep("depression.*condition_occurrence_id", current_cols, value = TRUE)
cat("Depression-related columns:\n")
print(depression_cols)
```

# Exercise 6: Converting OMOP IDs to Boolean Variables

Here's how we transform the condition IDs into analyzable format:

```{r exercise6}
cat("\n\n==================================================\n")
cat("EXERCISE 6: Transform condition IDs to analyzable format\n")
cat("==================================================\n\n")

# Function to convert OMOP IDs to boolean variables
convert_to_boolean <- function(table, variable_name, id_type, conns) {
  cat(sprintf("Converting %s to boolean...\n", variable_name))
  
  # Step 1: Construct the full variable name
  full_variable_name <- paste0(table, "$", variable_name, ".", id_type)
  
  # Step 2: Convert to numeric (IDs are often stored as strings)
  new_numeric_name <- paste0(variable_name, "_numeric")
  ds.asNumeric(
    x.name = full_variable_name, 
    newobj = new_numeric_name, 
    datasources = conns
  )
  
  # Step 3: Convert to boolean (1 if ID exists, 0 if not)
  ds.Boole(
    V1 = new_numeric_name, 
    V2 = 0, 
    Boolean.operator = "!=", 
    numeric.output = TRUE, 
    na.assign = 0, 
    newobj = variable_name,
    datasources = conns
  )
}

# Convert antenatal depression to boolean
convert_to_boolean("data", "antenatal_depression", "condition_occurrence_id", conns)

# Convert postpartum depression to boolean
convert_to_boolean("data", "postpartum_depression", "condition_occurrence_id", conns)

# Check the distribution of these new variables
cat("\nAntenatal depression prevalence:\n")
ds.table("antenatal_depression")
cat("\nPostpartum depression prevalence:\n")
ds.table("postpartum_depression")
```

# Exercise 7: Understanding Depression Patterns

Let's explore the depression patterns in our data.

```{r exercise7}
cat("\n\n==================================================\n")
cat("EXERCISE 7: Understanding Depression Patterns\n")
cat("==================================================\n\n")

# Create a histogram of Edinburgh scores
ds.assign(
  toAssign = "data$edinburgh_postnatal_depression_scale.value_as_number",
  newobj = "edinburgh_score",
  datasources = conns
)

cat("Edinburgh score distribution:\n")
ds.histogram("edinburgh_score", datasources = conns)

# Calculate summary statistics for Edinburgh scores
cat("\nEdinburgh score statistics:\n")
cat("Mean:", ds.mean("edinburgh_score", datasources = conns)$Global.Mean[1], "\n")
cat("Variance:", ds.var("edinburgh_score", datasources = conns)$Global.Variance[1], "\n")
cat("Quantiles:\n")
print(ds.quantileMean("edinburgh_score", datasources = conns))

# Create a cross-tabulation of depression types
cat("\nCross-tabulation of depression types:\n")
ds.table(rvar = "antenatal_depression", cvar = "postpartum_depression", datasources = conns)
```

# Exercise 8: Correlation Analysis

Let's explore correlations between variables.

```{r exercise8}
cat("\n\n==================================================\n")
cat("EXERCISE 8: Explore correlations\n")
cat("==================================================\n\n")

# Create simplified variable names
ds.assign("data$employment.value_as_concept_id", "employment", conns)
ds.assign("data$number_in_household.value_as_concept_id", "household_size", conns)

# Create a dataset for correlation
variable_list <- c("edinburgh_score", "age", "antenatal_depression", "postpartum_depression")
ds.dataFrame(
  x = variable_list,
  newobj = "cor_data",
  datasources = conns
)

# Calculate the correlation matrix
cat("Correlation matrix:\n")
cor_matrix <- ds.cor("cor_data", datasources = conns)
print(cor_matrix)
```

# Exercise 9: Building Statistical Models

Now let's test some hypotheses with statistical models.

```{r exercise9}
cat("\n\n==================================================\n")
cat("EXERCISE 9: Create and interpret GLM models\n")
cat("==================================================\n\n")

# Prepare all variables
ds.assign("data$marital_status.value_as_concept_id", "marital_status", conns)
ds.assign("data$religious_affiliation.value_as_concept_id", "religion", conns)

# Create modeling dataset
model_vars <- c(
  "edinburgh_score",
  "age",
  "employment",
  "marital_status",
  "household_size",
  "religion",
  "antenatal_depression",
  "postpartum_depression"
)

ds.cbind(
  x = model_vars,
  DataSHIELD.checks = FALSE,
  newobj = "model_data",
  datasources = conns
)

# Model 1: Simple linear regression
cat("Model 1: Simple linear regression\n")
cat("Research question: Do age and employment affect depression severity?\n\n")

model1 <- ds.glm(
  formula = "edinburgh_score ~ age + employment",
  data = "model_data",
  family = "gaussian",
  datasources = conns
)

cat("Model 1 Results:\n")
print(model1$coefficients[, c("Estimate", "Std. Error", "p-value")])

# Model 2: Logistic regression for postpartum depression
cat("\n\nModel 2: Logistic regression\n")
cat("Research question: What predicts postpartum depression?\n\n")

model2 <- ds.glm(
  formula = "postpartum_depression ~ antenatal_depression + edinburgh_score + age",
  data = "model_data",
  family = "binomial",
  datasources = conns
)

cat("Model 2 Results (with Odds Ratios):\n")
coef2 <- as.data.frame(model2$coefficients)
print(coef2[, c("Estimate", "P_OR", "low0.95CI.P_OR", "high0.95CI.P_OR", "p-value")])

# Model 3: Comprehensive model
cat("\n\nModel 3: Comprehensive model\n")
cat("Research question: How do social factors influence depression severity?\n\n")

model3 <- ds.glm(
  formula = "edinburgh_score ~ age + employment + marital_status + household_size + antenatal_depression + postpartum_depression",
  data = "model_data",
  family = "gaussian",
  datasources = conns
)

cat("Model 3 Results:\n")
coef3 <- as.data.frame(model3$coefficients)
significant <- coef3[coef3$`p-value` < 0.05, ]
if(nrow(significant) > 0) {
  cat("\nSignificant predictors (p < 0.05):\n")
  print(significant[, c("Estimate", "Std. Error", "p-value")])
} else {
  cat("\nNo significant predictors at p < 0.05\n")
}
```

# Exercise 10: Advanced Analysis - Interaction Effects

Let's test for interaction effects between variables.

```{r exercise10}
cat("\n\n==================================================\n")
cat("EXERCISE 10: Test for interaction effects\n")
cat("==================================================\n\n")

cat("Testing interaction: Does employment modify the effect of antenatal depression?\n\n")

model4 <- ds.glm(
  formula = "edinburgh_score ~ antenatal_depression * employment + age",
  data = "model_data",
  family = "gaussian",
  datasources = conns
)

# Look for interaction terms
coef4 <- as.data.frame(model4$coefficients)
interaction_terms <- grep(":", rownames(coef4), value = TRUE)
if(length(interaction_terms) > 0) {
  cat("Interaction effects:\n")
  print(coef4[interaction_terms, c("Estimate", "Std. Error", "p-value")])
}
```

# Clean Up

Always remember to log out from the servers!

```{r cleanup}
# Very important!!!
datashield.logout(conns)
```

# Summary

In this tutorial, you've learned how to:

1. **Connect to OMOP data** through DataSHIELD
2. **Discover the nature of a dataset** through systematic exploration
3. **Transform OMOP concepts** into analyzable variables using boolean conversion
4. **Perform statistical analyses** while preserving privacy
5. **Build and interpret models** with federated data

## Key Findings from Our Analysis

Through exploration, we discovered this was a **maternal mental health study** from APHRC with:
- 188 female participants of reproductive age
- 38% experiencing antenatal depression
- 17% experiencing postpartum depression
- Strong correlation between Edinburgh scores and depression diagnoses
- Complex relationships between social factors and mental health outcomes

## Next Steps

Consider exploring:
- Time-based analyses if longitudinal data is available
- More complex interaction effects
- Machine learning approaches with DataSHIELD
- Integration with other OMOP databases for meta-analysis

Remember: The power of DataSHIELD lies in its ability to analyze sensitive health data while preserving privacy!
